Bayesian exponential random graph modelling of KWL Burial network

### 2020-04-07 analysis/paper/paper.Rmd
add code to the rmd file and edit the method section

@@

We use burial data from different time periods at Kiwulan and consider the effects of both exogenous and endogenous factors that may provide meaningful explanation of the tendency of burials correlate with each other and the underlying mechnism.

n concerning connectivity, inter-cultural contacts and exchange among those sites. Our purpose is to understand the characteristics of burial network and the mechanism of the relationship influence on the pattern of accumulation of wealth during different foreign contact periods. Previous studies have shown that there are some evidence of social inequality observed from Kiwulan but the degree over time and causes remain unclear. To answer our question for the shifts in social inequality during and after the arrival of the foreign presences, we hypothesize:

@@

2. gwesp (geometrically weighted edgewise shared partners): triad-based clustering based on a dyad-based configurations (not node-based) that counts the number of times that both nodes have a tie to a third node. It is a measure of the tendency for two actors who are tied to have x partners in common. In orther words, a tendency for those with shared partners to become tied, or tendency of ties to cluster together.

3. gwdegree (geometrically weighted degree distribution): a parametric form that represents the frequency distribution for nodal degree, each node counts only once. The local configuration is degrees and stars that are equivalent representations for the distribution of node-based edge counts. A measure of the tendency of the number of edges into a node. gwdegree estimates the change in tie likelihood given the degree of the nodes involved, but with marginally decreasing weighting as degree increases(https://www.r-bloggers.com/ergm-tutorial/). This statistic captures the tendency towards centralisation in the degree distribution of the network.

@@

```{r pre-model-ergm-1}
model.ergm <- burial_network_pre ~ 
  edges + # ties, a measure of density, equal to kstar(1) for undirected networks
  density +
  gwdegree(0.5, fixed = TRUE) +
  triangle # triad relation, a measure of clustering or cohesion, also called transitive triple in undirected network, can be replaced by "gwdegree"
summary(model.ergm)
```

```{r pre-model-ergm-2}
# check out the terms: http://mailman13.u.washington.edu/pipermail/statnet_help/2010/000575.html
# also check here Morris et al. (2008)
model.2 <- burial_network_pre ~ edges + # density
  gwesp(0.2, fixed = TRUE) +  # transitivity(cohesion; triangle)
  # number means weight parameter alpha, which controls the rate of declining marginal returns
  # fixed = TRUE means the scale parameter lambda is fit as a curved exponential-family model
  # not much difference in a range of 0-1.5, the lower the value of the scaling parameter, the less likely the model is to be degenerate
  # ergm can estimate the parameter from the data by using fixed=FALSE

  gwdegree(0.8, fixed = TRUE)  # popularity(degree; star), the frequency distribution for nodal degrees
  # tendency of being in contact with multiple partners, measures of centralisation
  # distribution of node-based edge counts, each node counts only once
  # number means weight parameter decay
  # The closer decay is to zero, the more gwdegree considers low degree nodes relative to high degree nodes
summary(model.2)
```

```{r pre-model-ergm-3}
model.3 <- burial_network_pre ~ edges +  # the overall density of the network
  nodematch('quantity') + # quantity-based homophily, categorical nodal attribute
  # the similarity of connected nodes
  gwesp(0.2, fixed = TRUE) +  # transitivity
  gwdegree(0.8, fixed = TRUE) # popularity
summary(model.3)
```

@@

```{r pre-model-bergm}
# Specify a prior distribution: normal distribution (low density and high transitivity)
prior.mean <- c(-3, 0, 1, 0) # prior mean corresponds to mean for each parameter
# follow Alberto Caimo et al. (2015) hospital example
prior.sigma <- diag(3, 4, 4) # covariance matrix structure
# where the dimension d corresponds to the number of parameters, 𝜇 is mean vector and Σprior is a d × d covariance matrix.

# Estimated posterior means, medians and 95% credible intervals for Models.3
# bergmM: Bayesian exponential random graphs models under missing data using the approximate exchange algorithm
parpost <- bergm(model.3,
                  prior.mean  = prior.mean,
                  prior.sigma = prior.sigma,
                  burn.in     = 200, # burn-in iterations for every chain of the populationm, drops the first 200
                  main.iters  = 2000, # iterations for every chain of the population
                  aux.iters   = 10000, # MCMC steps used for network simulation
                  nchains     = 8, # number of chains of the population MCMC
                  gamma       = 0.7) # scalar; parallel adaptive direction sampling move factor, acceptance rate

summary(parpost) 
```

@@

>>>>

```{r read-data}
library(readxl)
library(tidyverse)
library(here)

# read data
burial <- read_excel(here("analysis", "data", "raw_data", "Kiwulan_Burials.xlsx"))
```

```{r tidy-data}
# combine some periods
burial_three_period_tidy <-
  burial %>%
  rename(burial_label = ID) %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>%
  mutate(Gold_leaf = ifelse(Gold_leaf == "shatter", "1", Gold_leaf),
         Stoneware = ifelse(Stoneware == "base", "1", Stoneware),
         Stamped_ceramic = ifelse(Stamped_ceramic == "cluster", "1", Stamped_ceramic)) %>%
  mutate_at(21:ncol(.), as.numeric) %>%
  janitor::remove_empty(which = "cols") %>%
  mutate(total = rowSums(.[c(21:48, 50, 55, 56)], na.rm = TRUE)) %>% #prestige goods
  mutate(Porcelain = rowSums(.[c(40:48, 55, 56)], na.rm = TRUE)) %>% #B&W, porcelain, Anping, kendi
  mutate(Porcelain = ifelse(Porcelain == 0, NA, Porcelain)) %>%
  mutate(quantity = case_when(
    total == 0 ~ "none",
    total > 0 & total <= 7 ~ "low",
    total > 7 & total < 100 ~ "medium",
    total >= 100 ~ "high",
    TRUE ~ "other")) %>% # the classification is based on the result of histogram
  mutate(Age_scale = case_when(
    `Age` %in% c("1","2") ~ "0-12",
    `Age` == "3" ~ "12~20",
    `Age` %in% c("4","5","6","7","8") ~ "+20",
    TRUE ~ "NA")) %>%
  mutate(Gold_bead_low = ifelse(Golden_bead == 1, 1, NA),
         Gold_bead_med = ifelse(Golden_bead > 1 & Golden_bead <10, 1, NA),
         Gold_bead_high = ifelse(Golden_bead > 10, 1, NA),
         Agate_bead_low = ifelse(Agate_bead == 1, 1, NA),
         Agate_bead_med = ifelse(Agate_bead > 1 & Agate_bead <10, 1, NA),
         Agate_bead_high = ifelse(Agate_bead > 10, 1, NA),
         `Indo-Pacific_bead_low` = ifelse(`Indo-Pacific_bead` < 100, 1, NA),
         `Indo-Pacific_bead_med` = ifelse(`Indo-Pacific_bead` > 100 & `Indo-Pacific_bead` < 900, 1, NA),
         `Indo-Pacific_bead_high` = ifelse(`Indo-Pacific_bead` > 900, 1, NA)) %>% 
  # the low-med-high index is determined by interpretation for histograms for those burial good types
  select(burial_label,
         Phase,
         Age_scale,
         Gold_bead_low,
         Gold_bead_med,
         Gold_bead_high,
         Agate_bead_low,
         Agate_bead_med,
         Agate_bead_high,
         #Agate_bead, #female burials
         #Golden_bead,
         Porcelain,
         Gold_leaf, #prestige good
         fish_shape_knit, #prestige good
         #Bell, #children's burials
         quantity)
         #total) # select specific variable to drop columns (uninformative variables)

# number of each phase
burial_three_period_number <-
  burial_three_period_tidy %>%
  count(Phase)
```

```{r network-object-pre}
# filter pre burials
burial_pre <-
  burial_three_period_tidy %>%
  filter(Phase == "pre") %>%
  janitor::remove_empty(which = "cols")

# create node list using burial index by phase
#--------------------pre--------------------------------
nodes_pre <-
  burial_three_period_tidy %>%
  filter(Phase == "pre") %>%
  select(burial_label) %>%
  rowid_to_column("id")

# pair wise combinations for burials as index for later map function
library(gtools)
burial_comb_pre = combinations(length(nodes_pre$burial_label), #t(combn(nodes_pre$burial_label, 2))
                               2, nodes_pre$burial_label,
                               repeats = TRUE)
colnames(burial_comb_pre) = c("burial_1", "burial_2")
burial_comb_pre = as_tibble(burial_comb_pre)

# create list for each burial that contains the burial good types and their counts
edge_list_pre <-
  burial_three_period_tidy %>%
  select(burial_label, 4:12) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  #mutate(burial_connection = rep(unique(burial_label), length.out = length(burial_label)))
  group_by(burial_label) %>%
  nest()

# pair-wise list
common_counts_lst_pre <-
map2(burial_comb_pre$burial_1,
     burial_comb_pre$burial_2,
     ~bind_cols(
       edge_list_pre %>%
       filter(burial_label == .x) %>%
       unnest(data),
       edge_list_pre %>%
         filter(burial_label == .y) %>%
         unnest(data)) %>%
       rowwise() %>%
       mutate(common_counts = sum(count, count1)))

# count of ornament types in common between each pair of burials
common_counts_vct_pre <- map_int(common_counts_lst_pre, ~sum(!is.na(.x$common_counts)))

burial_comb_with_common_counts_pre <-
burial_comb_pre %>%
  mutate(common_counts = common_counts_vct_pre) %>%
  mutate(common_counts = ifelse(burial_1 == burial_2, 0, common_counts)) # use 0 for self loop

# change label to ids for node linking
edges_pre <-
  burial_comb_with_common_counts_pre %>%
  left_join(nodes_pre, by = c("burial_1" = "burial_label")) %>%
  rename(from = id) %>%
  left_join(nodes_pre, by = c("burial_2" = "burial_label")) %>%
  rename(to = id)

edges_for_network_pre <-
  select(edges_pre, from, to, common_counts) %>%
  filter (!common_counts == 0) # remove rows with no goods in common
 # %>% mutate(common_counts = ifelse(common_counts > 1, 1, common_counts)) # for unweighted network
```

@@

We use burial data from different time periods at Kiwulan and consider the effects of both exogenous and endogenous factors that could provide meaningful explanation of the characteristics of burial network and tendency of burials correlate with each other. To be specific, our purpose is to understand the underlying mechanisms of the network structures and how the relationship influence on the pattern of accumulation of wealth during different foreign contact periods. We focus on connectivity, clusters, centralization, and homophily among those burials that could reflect the social relations between individuals guided by social structures.Previous studies have shown that there are some evidence of social inequality observed from Kiwulan but the degree over time and causes remain unclear. To answer our question for the shifts in social inequality to a more wealth differentiation during and after the arrival of the foreign presences, we hypothesize:

@@

2. gwesp (geometrically weighted edgewise shared partners): triad-based clustering based on a dyad-based configurations (not node-based) that counts the number of times that both nodes have a tie to a third node. It is a measure of the tendency for two actors who are tied to have x partners in common. In other words, a tendency for those with shared partners to become tied, or tendency of ties to cluster together.

3. gwdegree (geometrically weighted degree distribution): a parametric form that represents the frequency distribution for nodal degree, each node counts only once. The local configuration is degrees and stars that are equivalent representations for the distribution of node-based edge counts. A measure of the tendency of the number of edges into a node. This statistic captures the tendency towards centralisation in the degree distribution of the network.

```{r network-ggraph}
# network analysis using ggraph pkg
library(tidygraph)
library(ggraph)

relation_tidy_pre <- tbl_graph(nodes = nodes_pre,
                               edges = edges_for_network_pre,
                               directed = FALSE)

relation_tidy_pre %>%
  activate(edges) %>%
  arrange(desc(common_counts))

ggraph(relation_tidy_pre) +
  geom_edge_link() +
  geom_node_point() +
  theme_graph()

ggraph(relation_tidy_pre, layout = "graphopt") +
  geom_node_point() +
  geom_edge_link(aes(width = common_counts), alpha = 0.8) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_text(aes(label = burial_label), repel = TRUE) +
  labs(edge_width = "common item") +
  theme_graph()
```

```{r network-network-pkg}
# network analysis using network pkg
library(network)

burial_network_pre <-
  network(edges_for_network_pre, # the network object
          vertex.attr = nodes_pre, # node list
          directed = FALSE, # specify whether the network is directed
          ignore.eval = FALSE, # FALSE = weighted
          loops = FALSE, # do we allow self ties (should not allow them)
          matrix.type = "edgelist") # the type of input

# plot
plot(burial_network_pre, vertex.cex = 1) # last two nodes are missing?? only 27 nodes
```

```{r add-node-attribute-color}
# prepare notwork objects for Bayesian ERGMs 
library(statnet)
library(Bergm)

# Add attributes to the network object burial_network_pre
set.vertex.attribute(burial_network_pre, "quantity", burial_pre$quantity)
set.vertex.attribute(burial_network_pre, "age", burial_pre$Age_scale)

# plot
set.seed(30)
quantity <- get.vertex.attribute(burial_network_pre, "quantity")
age <- get.vertex.attribute(burial_network_pre, "age")

ID <- get.vertex.attribute(burial_network_pre, "burial_label") # not sure how to get id on the network plot

plot(burial_network_pre,
     vertex.col = "quantity", # age
     vertex.cex = 1.5)

legend("topleft",
       col = c(2, 3, 1, 4), # need to adjust each time
       pch    = 20,
       legend = unique(quantity), # age
       title  = 'Burial good counts')

#? can't get the items in legend in order, tried as.factor and set their level but does not work
#? can't match color with categories, using an odd method at the moment
```

@@

```{r network-ERGMs}
# model 1
model.ergm <- burial_network_pre ~
  edges + # ties, a measure of density, equal to kstar(1) for undirected networks
  density +
  triangle # triad relation, a measure of clustering or cohesion, also called transitive triple in undirected network, can be replaced by "gwdegree"
summary(model.ergm)

# model 2, check here Morris et al. (2008) 
# check out the terms: http://mailman13.u.washington.edu/pipermail/statnet_help/2010/000575.html
model.2 <- burial_network_pre ~ edges + # density
  gwesp(0.2, fixed = TRUE) +  # transitivity(cohesion; triangle)
  # number means weight parameter alpha, which controls the rate of declining marginal returns
  # fixed = TRUE means the scale parameter lambda is fit as a curved exponential-family model
  # not much difference in a range of 0-1.5, the lower the value of the scaling parameter, the less likely the model is to be degenerate
  # ergm can estimate the parameter from the data by using fixed=FALSE

  gwdegree(0.8, fixed = TRUE)  # popularity(degree; star), the frequency distribution for nodal degrees
  # tendency of being in contact with multiple partners, measures of centralisation
  # distribution of node-based edge counts, each node counts only once
  # number means weight parameter decay
  # The closer decay is to zero, the more gwdegree considers low degree nodes relative to high degree nodes
summary(model.2)

# model 3 with node attributes
model.3 <- burial_network_pre ~ edges +  # the overall density of the network
  nodematch('quantity') + # quantity-based homophily, categorical nodal attribute, the similarity of connected nodes
  nodematch('age') +
  gwesp(0.2, fixed = TRUE) +    # transitivity
  gwdegree(0.8, fixed = TRUE)   # popularity
summary(model.3)
```

```{r Bayesian-ERGMs}
# Specify a prior distribution: normal distribution (low density and high transitivity)
prior.mean <- c(-3, 0, 0, 1, 0) # prior mean corresponds to mean for each parameter
# follow Alberto Caimo et al. (2015) hospital example
prior.sigma <- diag(3, 5, 5) # covariance matrix structure
# where the dimension d corresponds to the number of parameters, 𝜇 is mean vector and Σprior is a d × d covariance matrix.

# Estimated posterior means, medians and 95% credible intervals for Models.3
# bergmM: Bayesian exponential random graphs models under missing data using the approximate exchange algorithm. Use this because of missing data for age variable
parpost <- bergmM(model.3,
                  prior.mean  = prior.mean,
                  prior.sigma = prior.sigma,
                  burn.in     = 200, # burn-in iterations for every chain of the population, drops the first 200
                  main.iters  = 2000, # iterations for every chain of the population
                  aux.iters   = 10000, # MCMC steps used for network simulation
                  nchains     = 8, # number of chains of the population MCMC
                  gamma       = 0.7) # scalar; parallel adaptive direction sampling move factor, acceptance rate

# summary of parameter estimation
summary(parpost)

# plot the parameter posterior distribution
plot(parpost)
```

@@

@@

```{r Bayesian-gof}
# Model assessment, Bayesian goodness of fit diagnostics:
bgof(parpost,
     aux.iters = 10000,
     n.deg     = 14,
     n.dist    = 15,
     n.esp     = 9)

# An estimated ERGM is fitting perfectly a certain observed network if the red line falls inside this interval
# a result that is very difficult to obtain in practice (Caimo 2017)
```
